---
title: "MERL Tech DC workshop: Up your Data Game with R"
author: "Charles Guedenet"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
        toc_depth: 2
        number_sections: true
---

This tutorial was developed for a workshop at MERL Tech DC in 2018 and presented with Jonathan Seiden. You can find the R code and slides on [Github](https://github.com/cguedenet/MERL-Tech-workshop.git).   
  
The code and examples below are an introduction to the possibilities of using the free and open-source R programming language and software environment for data processing, analysis, and visualization.  
  
#Introduction
This was built in an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code. 

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
# activate the packages you have already installed using the library() function
library(prettydoc)
library(sjmisc) # useful for data wrangling work like recoding, grouping variables, summarizing data
library(knitr) # for creating reports in Word, pdf, html
library(ggthemes) # to add themes to ggplots 
library(corrr) # to create correlation matrices and plots
library(tidyverse) #includes a set of packages that work well together (ggplot, tidyr, dplyr, and others)
library(psych) # for summarizing data
library(forcats) # useful for working with factor variabls
library(RCurl) # for reading in web data or files
```

```{r import data, message=FALSE, warning=FALSE, echo=FALSE}
# read in your data
df <-read.csv(text=getURL("https://raw.githubusercontent.com/cguedenet/guedenet-website/master/2016.coders.survey.csv"))
```

R Studio is an "Integrated Development Environment", or IDE. This means it is a front-end for R that makes it much easier to work with. R Studio is also free, and available for Windows, Mac, and Linux platforms.
```{r Rstudio intro, message=FALSE, warning=FALSE, echo=FALSE, eval = FALSE}

# This is where you execute R ccommands. R interprets what you write and prints the results
1 + 1

# A basic concept in programming is called a variable. A variable allows you to store a value (e.g. 5) or an object (e.g. a function) in R for easy access later. 
## TIP: use "Alt" + "-" instead of "<" + "-"

#creating variables for later use. Use the less sign followed by a dash to create a variable

width <- 5                       # assigns the value 5 to the variable "width"
length <- 8                      # assigns the value 8 to the variable "width"
area <- width * length           # multiple the values of both previous variables to create a new variable: "area" 

area  # print the result
```

#Cleaning your data  
Analyzing survey data typically starts with cleaning, recoding, and restructuring, or even joining data sets. For example, you may want to know how many missing cases there are or how many people responded to each question. Or you may want to group certain continuous variables like ages or income into ranges. Lastly, you may want to find and deal with outliers.  
  
Functions like distinct() can quickly get rid of duplicate rows across all variables or for specific variables (e.g. a person's ID or email address)  
    
Functions like mutate() can help create new variables. For example, creating a new variable for age ranges based on an existing variable for age.  
   
```{r cleaning, message=FALSE, warning=FALSE, echo=FALSE}

# remove duplicated rows (if any). The distinct() function keeps only unique rows.
df <- distinct(df)

# You can also choose to remove duplicated rows for specific variables like ID.x and ID.y
df <- distinct(df, ID.x, ID.y, .keep_all = TRUE)

# Group responses and create a new variable using mutate function (dplyr package)
#this creates a new variable that recodes the Age variable into 5 age categories 

df <- df %>% mutate(AgeCut = cut(Age, c(10,29,34,50,65,100)))
```

#Overview of survey data and basic analysis  
When you're analyzing survey data, one of the first things you need to do is get an overview of your data. For example, you may want to know basic stats for continuous variables or frequency tables for other types of data.  

##Get summary stats  
```{r summary, warning=FALSE, message=FALSE ,echo=FALSE, results = 'asis'}
summary(df$Age)
```

##Add new stats to your table  
```{r summary2,warning=FALSE, message=FALSE ,echo=FALSE, results = 'asis'}
df %>% summarise(
  mean = mean(Age, na.rm = TRUE),
  median = median(Age, na.rm = TRUE),
  IQR = IQR(Age, na.rm = TRUE),
  n = n()
  ) %>%
  kable()
```
  
##Add a grouping variable
```{r grouping, warning=FALSE, message=FALSE ,echo=FALSE, results = 'asis'}
df %>%
  group_by(Gender) %>%
  summarise(
    mean = mean(Age, na.rm = TRUE),
    median = median(Age, na.rm = TRUE),
    IQR = IQR(Age, na.rm = TRUE), n = n()
    ) %>%
  kable()
```

###Exercise: 
1. Get summary statistics for another continuous variable, like income, MonthsProgramming, ExpectedEarning, MoneyForLearning, etc. (reference the df object)
   
2. Add or change the summary stats you want to calculate (Other useful functions: mean, median, sd, IQR, min, max, quantile, first, last, nth, n, n_distinct)
   
3. Add a new grouping variable
  
##Quickly summarize data
Here's another way to quickly generate summary statistics for select numeric variables using the psyche package  
```{r warning=FALSE, message=FALSE ,echo=FALSE, eval=TRUE, results = 'asis'}
df %>% select(c("Age", "Income", "ExpectedEarning", "HoursLearning", "Gender"))%>%
  describeBy() %>% kable()
```

#Working with Categorical data  
    
##Summarize categorical data     
  
Summarize categorical data and create frequency tables using sjmisc package    
```{r warning=FALSE, message=TRUE ,echo=TRUE, eval=TRUE, results = 'asis'}
# summarize categorical data by creating frequency tables using sjmisc package. Frequency table for the variable: "SchoolDegree"
df %>% frq(SchoolDegree) %>% kable()
```

##Create cross-tabulations with two or more variables
```{r warning=FALSE, message=FALSE ,echo=FALSE, eval=TRUE, results = 'asis'}
df %>% flat_table(SchoolDegree, Gender) %>%
  data.frame() %>%
  spread(Gender,Freq) %>%
  kable()
```

##Create marginal tables using "row", "col", or "cell"
```{r warning=FALSE, message=FALSE ,echo=FALSE, eval=TRUE, results = 'asis'}
df %>% flat_table(SchoolDegree, Gender, margin = "col") %>%
  data.frame() %>%
  spread(Gender,Freq) %>%
  kable()
```
   
###Exercise:   
1. Create other cross-tabulations by changing the variables   
2. try changing the margin argument to col, row, or cell
    
     
#Create charts using ggplot2   
   
##create a simple column chart
```{r warning=FALSE, message=FALSE ,echo=FALSE, eval=TRUE, results = 'asis'}
df %>% ggplot(aes(EmploymentField)) + geom_bar()
```
   
####flip it and order bars
```{r warning=FALSE, message=FALSE ,echo=TRUE, eval=TRUE, results = 'asis'}

df %>% ggplot(aes(fct_infreq(EmploymentField))) +
  geom_bar() +
  coord_flip()
```
   
####Get rid of NA
```{r warning=FALSE, message=FALSE ,echo=TRUE, eval=TRUE, results = 'asis'}
df %>% filter(EmploymentField!="") %>%
  ggplot(aes(fct_infreq(EmploymentField))) +
  geom_bar() +
  coord_flip()
```
   
##Create a simple histogram  
```{r warning=FALSE, message=FALSE ,echo=TRUE, eval=TRUE, results = 'asis'}
df %>% ggplot(aes(x=Age)) + 
  geom_histogram()
```
    
####Adding titles and themes to charts  
```{r warning=FALSE, message=FALSE ,echo=TRUE, eval=TRUE, results = 'asis'}
df %>% ggplot(aes(x=Age)) +
  geom_histogram(color = "white") +
  theme_minimal() +
  labs(title = "My Title", subtitle = "My subtitle") 
```
      
Try another theme  
```{r warning=FALSE, message=FALSE ,echo=FALSE, eval=TRUE, results = 'asis'}
df %>% ggplot(aes(x=Age)) +
  geom_histogram(color = "white", bins = 35) +
  theme_fivethirtyeight() +
  labs(title = "My Title", subtitle = "My subtitle") 
```
   
###Exercise:   
1 Change the x variable with another numeric variable   
2 Change the theme & titles   
3 Change the number of bins   
    
   
##Create a density plot   
```{r density plot, message=FALSE, warning=FALSE, echo=TRUE}
# Compare only males and females comparison
df %>% filter(Gender == "female" | Gender == "male") %>% 
  ggplot(aes(x=Age, fill = Gender)) + 
  geom_density(alpha = .6) + 
  theme_fivethirtyeight()
```

##Other examples of charts with the ggplot package    
Relationship between employment status and job preference   
```{r examples, message=FALSE, warning=FALSE, echo=TRUE}
df %>% flat_table(EmploymentStatus, JobPref) %>% data.frame() %>%
  ggplot(aes(EmploymentStatus, JobPref)) +
  geom_tile(aes(fill = Freq), colour = "white") +
  scale_fill_continuous() +
  coord_fixed(ratio = 1) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
```
   
###Stacked bar chart   
Comparing gender breakdown across ages
```{r warning=FALSE , message=FALSE ,echo=TRUE}
df %>% select(Age, Gender) %>%
  group_by(Age, Gender) %>%
  summarise(count=n()) %>%
  ggplot(aes(Age,count,fill=Gender)) +
  geom_bar(stat='identity', position='fill', color='white') + 
  xlim(c(10,70))
```